FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-26 ess_marmo.tex
Update on Overleaf.

\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay,
Orsay, France}
\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center,
Flagstaff, AZ, USA}
\affiliation{3}{LESIA, Observatoire de Paris, PSL Research University,
CNRS, Sorbonne Universit\'es, UPMC Univ. Paris 06, Univ. Paris Diderot,
Sorbonne Paris Cit\'e, Meudon, France}
\affiliation{4}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
\affiliation{5}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
\affiliation{6}{Jacobs University, Bremen, Germany}
\affiliation{7}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.). Planetary missions produce an impressively growing 
amount of diverse data requiring an evolution from a mostly manual-based visual 
analysis to a more automated and quantitative analysis.
Interoperability, openness of data formats, and shared processing techniques are 
a necessity to efficiently extract scientific information from the data and 
to guarantee the reproducibility of the scientific results.
Unfortunately, the technologies and data formats used by researchers for
planetary surface studies and the field of astronomy 
diverged as these related, but almost completely isolated domains evolved.  
% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
% 
% > Unfortunately, the technologies 
% > and data formats used by researchers for planetary surface studies and the field of astronomy 
% > diverged as these related, but almost completely isolated domains evolved.  
% 
% Is this too strong a statement?
% 
% ^.
In this paper we will describe how a small addition to the Flexible Image Transport System (FITS) 
standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
surface investigations. We will also show  how FITS metadata can easily be transformed in the PDS4 metadata
archival model and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from data acquisition to final visualization.
\end{abstract}

@@

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open, flexible, and largely implemented in open and efficient
processing tools allowing to handle large amounts of raster data.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary surface domain could help to
homogenize methods from acquisition to visualization, while
allowing for optimized data processing across the planetary and astronomy domains.

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition, transfer and
archiving of telescope data by astronomical observatories.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

@@

In addition, WCS resolves the ambiguity between east-positive and west-posited
longitude systems, as the pixel conversion in world coordinates is set by an
oriented matrix.
If needed, FITS does offer the possibility to fully describe the projected
coordinate system using several alternative WCS definitions (meters, degrees,
east and west longitude, etc.)\citep{greisencalaII}.

@@

The coordinate system must be defined with respect to a specific inertial reference frame.

Multiple coordinate system descriptions are allowed in the same header: this is a valuable
property for planetary surface imagery, as angular (`deg') and linear (`m') description of
the surface are often needed at the same time.
Further below 3D keyword will be proposed for FITS metadata, in order to standardize
the conversion between angular and linear coordinates: alternate description will still
be important in our framework as standard libraries (e.g.
wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
coordinate tracking more efficient.

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined (defaulted to \texttt{`ICRS'})}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
correspond to the target body. We propose to standardize those codes as listed in Table
\ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X, xPY as
Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in degrees
and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
attributes.
In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
\item{CUNITn.
Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
and to meters (m) for linear ones.}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space
Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from
solar system exploration missions.
Up to now it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and
can be integrated with other datasets thanks to its modular structure.
Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is
formatted as FITS, projected by interpolating the original observations to a grid regularly spaced
in planetocentric coordinates (i.e., latitude and longitude).
We used the planetary FITS standard described in this work in order to make the output file compliant
with the great part of the FITS readers, DS9 and JS9 above all.
The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end
of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the
tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based
environment.

@@

Aladin\footnote{http://aladin.u-strasbg.fr/} and its HiPS generator have been updated to support
several planetary conventions.
First, while the sky is displayed from the inside of the unit sphere, planetary
surfaces are observed -- and must be displayed -- from the exterior.
It means that the direction of the east on the horizontal axis has to be
inverted when displaying planetary data.
Then, planetary missions usually use different projections when providing
images of planets polar caps and equatorial region.
Often the former are in the stereographic (STG) projection while the later
is in the Plate Carr\'ee (CAR) projection. 
This specificity has been taken into account into the HiPS generator.
Also, in the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
For example, it will be able to take into account a planet's radii to compute
distances on the surface of the body.
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.
We expect all projections to be available in a forthcoming release.

@@

\section{Discussion and Perspectives}
\label{sec:disc}
FITS discussion in astronomy\citep{THOMAS2015133} 
let's astronomy learn from planetary sciences too: accent must be put on interoperability
.... blabla
Planetary surface is lacking open and general software for massive processing and can benefit
of astronomy experience. In addition to that WCS projection scheme is simpler than the
historical one used in planetary applications. 
Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
Reducing custom data format conversions during the processing chain but
allowing easy conversion for final products following the needed application.

>>>>

\affiliation{1}{GEOPS, Univ. Paris-Sud, CNRS, Univ. Paris-Saclay,
Orsay, France}
\affiliation{2}{U. S. Geological Survey, Astrogeology Science Center,
Flagstaff, AZ, USA}
\affiliation{3}{LESIA, Observatoire de Paris, Universit\'e PSL, CNRS, Sorbonne Universit\'e,
Univ. Paris Diderot, Sorbonne Paris Cit\'e, 5 place Jules Janssen, 92195 Meudon, France}
\affiliation{4}{ASI Science Data Center, c/o ASI, Via del Politecnico snc, 00133, Rome, Italy}
\affiliation{5}{INAF-OAR, Via Frascati n. 33, 00078, Monte Porzio Catone (RM), Italy}
\affiliation{6}{Jacobs University, Bremen, Germany}
\affiliation{7}{Observatoire astronomique de Strasbourg, Universit\'e de Strasbourg, CNRS, UMR 7550, 11 rue de l'Universit\'e, 67000, Strasbourg, France}

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.). Planetary missions produce an impressively growing 
amount of diverse data requiring an evolution from a mostly manual-based visual 
analysis to a more automated and quantitative analysis.
Interoperability, openness of data formats, and shared processing techniques are 
a necessity to efficiently extract scientific information from the data and 
to guarantee the reproducibility of the scientific results.
Unfortunately, the technologies and data formats used by researchers for
planetary surface studies and the field of astronomy 
diverged as these related, but almost completely isolated domains evolved.  
% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
% 
% > Unfortunately, the technologies 
% > and data formats used by researchers for planetary surface studies and the field of astronomy 
% > diverged as these related, but almost completely isolated domains evolved.  
% 
% Is this too strong a statement?
% 
% ^.
In this paper we will describe how a small addition to the Flexible Image Transport System (FITS) 
standard, widely used in the astronomy investigations, will allow FITS to be more easily used in planetary
surface investigations. We will also show  how FITS metadata can easily be transformed in the PDS4 metadata
archival model and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from raw data formatting to final visualization.
\end{abstract}

@@

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO), therefore making obvious its connection to the Planetary Virtual
Observatory initiative and the planetary Table Access Protocol (EPN-TAP)
\citep{ERARDepntap}.
FITS format is open, flexible, and largely implemented in open and efficient
processing tools allowing to handle large amounts of raster data.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary surface domain could help to
homogenize methods from raw formatting to visualization, while
allowing for optimized data processing across the planetary and astronomy domains.

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition, transfer and
archiving of telescope data by astronomical observatories.
It had been used data exchange and archiving from several orbital telescopes.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

@@

In addition, WCS resolves the ambiguity between east-positive and west-positive
longitude systems, as the pixel conversion in world coordinates is set by an
oriented matrix.
If needed, FITS does offer the possibility to fully describe the projected
coordinate system using several alternative WCS definitions (meters, degrees,
east and west longitude, etc.)\citep{greisencalaII}.

@@

The coordinate system must be defined with respect to a specific inertial reference frame.
% * <chiara.marmo@u-psud.fr> 2018-02-26T11:33:56.098Z:
% 
% Please be more specific about the inertial frame
% 
% ^.
Multiple coordinate system descriptions are allowed in the same header: this is a valuable
property for planetary surface imagery, as angular (`deg') and linear (`m') description of
the surface are often needed at the same time.
Further below 3D keyword will be proposed for FITS metadata, in order to standardize
the conversion between angular and linear coordinates: alternate description will still
be important in our framework as standard libraries (e.g.
wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
coordinate tracking more efficient.

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving and distribution
standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined (defaulted to \texttt{`ICRS'})}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
correspond to the target body. We propose to standardize those codes as listed in Table
\ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X, xPY as
Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in degrees
and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
attributes.
In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
\item{CUNITn.
Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
and to meters (m) for linear ones.}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space
Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from
solar system exploration missions.
Currently, it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and
can be integrated with other datasets thanks to its modular structure.
Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is
formatted as FITS, projected by interpolating the original observations to a grid regularly spaced
in planetocentric coordinates (i.e., latitude and longitude).
We used the planetary FITS standard described in this work in order to make the output file
compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end
of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the
tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based
environment.

@@

Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support several planetary conventions.
First, while the sky is displayed from the inside of the unit sphere, planetary
surfaces are observed -- and must be displayed -- from the exterior.
It means that the direction of the east on the horizontal axis has to be
inverted when displaying planetary data.
Then, planetary missions usually use different projections when providing
images of planets polar caps and equatorial region.
Often the former are in the stereographic (STG) projection while the later
is in the Plate Carr\'ee (CAR) projection. 
This specificity has been taken into account into the HiPS generator.
Also, in the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
For example, it will be able to take into account a planet's radii to compute
distances on the surface of the body.
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.
We expect all projections to be available in a forthcoming release.

@@

\section{Discussion and Perspectives}
\label{sec:disc}
FITS discussion in astronomy\citep{THOMAS2015133} 
let's astronomy learn from planetary sciences too: accent must be put on interoperability
.... blabla
Planetary surface is lacking open and general software for massive processing and can benefit
of astronomy experience. In addition to that WCS projection scheme is simpler than the
historical Earth-based one used in planetary applications. 
Archiving flexibility and data model are today approached by PDS4 and EPN-TAP efforts.
Reducing custom data format conversions during the processing chain but
allowing easy conversion for final products following the needed application.
